A recipe for highly reproducible enrichment analysis (and why we need it)

### 2024-02-29 manuscript/manuscript.Rmd
addressing reviewer points

---
title: "A recipe for extremely reproducible enrichment analysis (and why we need it)"
date: "`r Sys.Date()`"
output:
  rmdformats::robobook:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 5
    fig_height: 5
bibliography: references.bib
csl: plos-computational-biology.csl
---

@@

Unreliable and irreproducible research is a significant problem that wastes resources and risks
undermining the public perception of science.
Previous work has highlighted that published enrichment analyses frequently suffer from statistical
and reporting flaws.
We sought to determine whether this translates into irreproducibility by examining whether the
findings of 20 open-access articles published in 2019 describing enrichment analysis with the
popular DAVID suite were reproducible.
We find that only four articles exhibited a high degree of concordance, while seven exhibited major
discrepancies, which we mainly ascribe to deficiencies in methodological reporting.
As the tool version used is no longer available, all articles using this tool pre-2021 (~20,800 studies)
including this sample of 20 articles are no longer able to be reproduced with the original tools.
Based on this, we suggest that results from web-based tools without long-term preservation features should
not be included in scientific publications due to the threat of link decay and short reproducibility horizon.
Relying exclusively on webtools for analysis may also be in breach of institutional and funder data management mandates.
We advocate for the adoption of extremely reproducible research workflows, and we provide a detailed protocol
for how to achieve it using a combination of best practices including literate programming, version control,
containerisation, documentation and persistent sharing of data and software.

@@

In addition to shared code, data and containerisation, there are a few other ways researchers can make
their work more reproducible.
Thorough documentation makes the reproducibility process a great deal easier and more
accessible, especially to novice bioinformaticians.
Version control keeps track of code and documentation changes over time and manages the edits made by
multiple contributors.
Literate programming using R Markdown or Jupyter notebooks enables thorough documentation by embedding
free text explanations together with the code and outputs such as charts and tables.
Code and data can be linked. This means that the code "knows" the location of the data,
avoiding any "file not found" errors @Peng2011-ia.
Lastly, reanalysis can be made as simple as possible by creating a master script that executes analysis
for each part sequentially, making the entire manuscript reproducible by executing just one or a few
commands, and without any manual steps @Lasser2020-mm.
Taken together, computational reproducibility is a spectrum, and the *five pillars*
framework has been described as a way of achieving extreme reproducibility @Ziemann2023-dg.
Currently, bioinformatics work at this gold standard is rare, despite the availability of such tools for
over a decade.

@@

These observations highlight that pathway enrichment tools are essential in the interpretation of omics data.
However, there are concerns that inadvertent misuse of these methods leads to unreliable results.
These potential problems include (i) the lack of a correct background for over-representation tests,
(ii) lack of p-value correction for multiple testing and (iii) lack of reporting detail
[@Timmons2015-ex; @Reimand2019-uw; @Wijesooriya2022-li; @Zhao2023-ok].
Without comprehensive methodological documentation and reproducibility, it isn't possible to verify
the validity of the methods and results.
We are concerned about the reliability of published enrichment analyses, as web-based and
point-and-click tools appear to be far more popular than computer script-based methods for this type
of work @Wijesooriya2022-li.
Although web-based tools are convenient, they are a potential reproducibility concern.
Incomplete reporting of methods, options and parameters is common.
Algorithms and functional annotation sets undergo regular updates, and previous versions are
typically not made available, hampering later reproducibility.
In some cases, the entire web resource can become unavailable, and hence irreproducible due to
a phenomenon known as link decay @Hennessey2013-tj.

@@

A solution to this is to use containerisation to provide a complete scripted workflow with prescribed
links to research data, external databases and complete record of parameter selection.
When designed correctly, these containers encapsulate bioinformatics workflows and could ensure their
computational reproducibility for decades to come with a few simple commands.
There are multiple barriers to entry for such workflows including (i) the added difficulty compared to
existing point-and-click solutions; (ii) time poorness of researchers to learn new skills; and (iii)
a lack of step-by-step guides and templates aimed at non-expert bioinformaticians.

We address this firstly by providing working templates for Docker image and R Markdown scripts for
common transcriptome and enrichment analysis routines.
Secondly, we provide step-by-step written and video guides to help users customise these templates for
their own work, deploy the analysis, verify the results, and share the materials.

@@

barplot(unlist(lapply(l,length)),col="lightgray",
  xlab="reproducibility score",ylab="no. articles",
  ylim=c(0,11))

grid()
text(0.7,1:7-0.5,labels=score1)
text(1.9,1:9-0.5,labels=score2)
text(3.1,1:4-0.5,labels=score3)

@@

We have developed a step-by-step protocol for how to conduct an extremely reproducible
enrichment analysis, which is attached as a supplementary file (Suppl Info 1) and published to
protocols.io with a permissive license [@Ziemann2023-kz].
This protocol is prepared in accordance with the *five pillars* principles [@Ziemann2023-dg], and includes
version control with git and GitHub, compute environment control using Docker, literate programming
with R Markdown, persistent code and data sharing with Zenodo and documentation of all of these to
empower others to reproduce the workflow.
Central to this protocol is the provision of template Dockerfile and an R Markdown script, which
are intended to be remixed and altered by end users to suit their project's needs.
The general steps involved are shown in **Figure 2**.

@@

```bash
sudo apt update && sudo apt install docker.io -y  # 30 sec
sudo docker run --platform linux/amd64 mziemann/enrichment_recipe Rscript -e 'rmarkdown::render("example.Rmd")' # 150 sec
docker cp $(docker ps -aql):/enrichment_recipe/example.html . # instant
firefox example.html
```
<p style="text-align: center;">
*Box 1. Linux shell commands to reproduce a transcriptome and enrichment analysis using a Docker image.*
</p>

@@

## Concluding remarks

@@

The problem of bioinformatic resources becoming unavailable is not a new one, being thoroughly described a
decade ago [@Hennessey2013-tj].
We are worried that a reliance on web tools can leave researchers in violation of their data management
responsibilities, as preservation mandates include software as well as original research data
[@NHMRC2019-wu ; @Harvard_2023-cl].

With this, we must reconsider whether web-based tools are valid for scientific publications if they are at
risk of becoming unavailable without notice due to link decay.
Some progress is being made to make web-based point-and-click tools more reproducible.
The Galaxy, GenePattern and RNAlysis projects offer graphical tools to create, edit and import
workflows that enable reproducibility [@Kuehn2008-su; @Goecks2010-lq; @Teichman2023-fx].
The Degust differential expression suite is a graphical interface to popular
expression tools like edgeR and limma and allows users to download the computer script to facilitate
future non-interactive reproducibility @Powell2019-lv.
The ShinyGO web-based enrichment tool is a satisfactory GUI-based alternative to DAVID as it provides Docker image
downloads for all past versions [@Ge2020-qe], facilitating preservation and aiding future reproducibility.
Still, such tools leave documentation and reproducibility gaps which are resolved with the "extremely reproducible"
approach detailed here.

Although we hope that extremely reproducible enrichment analysis will become more widespread, we do not
anticipate any dramatic changes in researcher behaviour unless the motivating factors are addressed.
Currently, reward structures in academia favour career progression for individuals who publish many
articles as opposed to individuals who publish fewer but more rigorously conducted works.
Solutions to this could include mandates by funding agencies, institutions or journals [@Garijo2013-zb],
badging of articles that comply with best practices @Kidwell2016-pm, and recognition of research rigour
in track record evaluations.

>>>>

---
title: "A recipe for highly reproducible enrichment analysis (and why we need it)"
date: "`r Sys.Date()`"
output:
  rmdformats::robobook:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 5
    fig_height: 5
bibliography: references.bib
csl: plos-computational-biology.csl
---

@@

Irreproducible research is a significant problem that wastes resources and risks
undermining the public perception of science.
Previous work has highlighted that published enrichment analyses frequently suffer from statistical
and reporting flaws.
We sought to determine whether this translates into irreproducibility by examining whether the
findings of 20 open-access articles published in 2019 describing enrichment analysis with the
popular DAVID suite were reproducible.
We find that only four articles exhibited a high degree of concordance, while seven exhibited major
discrepancies, which we mainly ascribe to deficiencies in methodological reporting.
To help combat such issues, we provide a detailed beginner-friendly enrichment analysis protocol
exemplifying current best practices including literate programming, version control, containerisation,
documentation and persistent sharing of data and code.
Finally, we discuss how web-based tools without long-term preservation features might inadvertently put
researchers in breach of their data retention mandates.

@@

In addition to shared code, data and containerisation, there are a few other ways researchers can make
their work more reproducible.
Thorough documentation makes the reproducibility process a great deal easier and more
accessible, especially to novice bioinformaticians.
Version control keeps track of code and documentation changes over time and manages the edits made by
multiple contributors.
Literate programming using R Markdown or Jupyter notebooks enables thorough documentation by embedding
free text explanations together with the code and outputs such as charts and tables.
Code and data can be linked. This means that the code "knows" the location of the data,
avoiding any "file not found" errors @Peng2011-ia.
Lastly, reanalysis can be made as simple as possible by creating a "main" script that executes analysis
for each part sequentially, making the entire manuscript reproducible by executing just one or a few
commands, and without any manual steps @Lasser2020-mm.
Taken together, computational reproducibility is a spectrum, and the *five pillars*
framework has been described as a way of achieving a high degree of reproducibility @Ziemann2023-dg.
Currently, bioinformatics work at this gold standard is rare, despite the availability of such tools for
over a decade.

@@

These observations highlight that pathway enrichment tools are essential in the interpretation of omics data.
However, there are concerns that inadvertent misuse of these methods leads to invalid results.
These potential problems include (i) the lack of a correct background for over-representation tests,
(ii) lack of p-value correction for multiple testing and (iii) lack of reporting detail
[@Timmons2015-ex; @Reimand2019-uw; @Wijesooriya2022-li; @Zhao2023-ok].
Without comprehensive methodological documentation and reproducibility, it isn't possible to verify
the validity of the methods and results.
We are concerned about the validity of published enrichment analyses, as web-based and
point-and-click tools appear to be far more popular than computer script-based methods for this type
of work @Wijesooriya2022-li.
Although web-based tools are convenient, they are a potential reproducibility concern.
Incomplete reporting of methods, options and parameters is common.
Algorithms and functional annotation sets undergo regular updates, and previous versions are
typically not made available, hampering later reproducibility.
In some cases, the entire web resource can become unavailable, and hence irreproducible due to
a phenomenon known as link decay @Hennessey2013-tj.

@@

One potential solution to this is to use containerisation to provide a complete scripted workflow with prescribed
links to research data, external databases and complete record of parameter selection.
When designed correctly, these containers encapsulate bioinformatics workflows and could ensure their
computational reproducibility for decades to come with a few simple commands.
There are multiple barriers to entry for such workflows including (i) the added difficulty compared to
existing point-and-click solutions; (ii) time poorness of researchers to learn new skills; and (iii)
a lack of step-by-step guides and templates aimed at non-expert bioinformaticians.

We address this firstly by providing working example templates for Docker image and R Markdown scripts for
common transcriptome and enrichment analysis routines.
Secondly, we provide step-by-step written and video guides to help users customise these templates for
their own work, deploy the analysis, verify the results, and share the materials.

@@

barplot(unlist(lapply(l,length)),col="lightgray",
  xlab="Reproducibility score",ylab="no. articles",
  ylim=c(0,11))

#grid()
text(0.7,1:7-0.5,labels=score1)
text(1.9,1:9-0.5,labels=score2)
text(3.1,1:4-0.5,labels=score3)

@@

We have developed a step-by-step protocol for how to conduct a highly reproducible
enrichment analysis, which is attached as a supplementary file (Suppl Info 1) and published to
protocols.io with a permissive license [@Ziemann2023-kz].
This protocol is prepared in accordance with the *five pillars* principles [@Ziemann2023-dg], and includes
version control with git and GitHub, compute environment control using Docker, literate programming
with R Markdown, persistent code and data sharing with Zenodo and documentation of all of these to
empower others to reproduce the workflow.
Central to this protocol is the provision of template Dockerfile and an R Markdown script, which
are intended to be remixed and altered by end users to suit their project's needs.
The general steps involved are shown in **Figure 2**.

@@

```bash
sudo apt update && sudo apt install docker.io -y  # 30 sec
sudo docker run --platform linux/amd64 mziemann/enrichment_recipe Rscript -e 'rmarkdown::render("example.Rmd")' # 150 sec
docker cp $(docker ps -aql):/enrichment_recipe/example.html . # instant
firefox example.html
```
<p style="text-align: center;">
*Box 1. Linux shell commands to reproduce a transcriptome and enrichment analysis using a Docker image.
Ubuntu 22.04 was used.*
</p>

@@

## Discussion

@@

The problem of bioinformatic resources becoming unavailable is not a new one, being thoroughly described a
decade ago [@Hennessey2013-tj].
We are worried that a reliance on web tools that lack open source code or containers can leave
researchers in violation of their data management responsibilities, as preservation mandates
include software as well as original research data
[@NHMRC2019-wu ; @Harvard_2023-cl].

With this, we must reconsider whether web-based tools are valid for scientific publications if they are at
risk of becoming unavailable without notice due to link decay.
Some progress is being made to make web-based point-and-click tools more reproducible.
The Galaxy, GenePattern and RNAlysis projects offer graphical tools to create, edit and import
workflows that enable reproducibility [@Kuehn2008-su; @Goecks2010-lq; @Teichman2023-fx].
The Degust differential expression suite is a graphical interface to popular
expression tools like edgeR and limma and allows users to download the computer script to facilitate
future non-interactive reproducibility @Powell2019-lv.
The ShinyGO web-based enrichment tool is a satisfactory GUI-based alternative to DAVID as it provides Docker image
downloads for all past versions [@Ge2020-qe], facilitating preservation and aiding future reproducibility.
Still, such tools leave documentation and reproducibility gaps which are resolved with the highly reproducible
approach detailed here.

Although the recipe described here is designed for gene expression analysis, it is modifiable to accommodate
other omics, for example genomic, epigenomic and proteomic data, given the names of analytes measured are
consistent with the annotation databases.

Although we hope that highly reproducible enrichment analysis will become more widespread, we do not
anticipate any dramatic changes in researcher behaviour unless the motivating factors are addressed.
Currently, reward structures in academia favour career progression for individuals who publish many
articles as opposed to individuals who publish fewer but more rigorously conducted works.
Solutions to this could include mandates by funding agencies, institutions or journals [@Garijo2013-zb],
badging of articles that comply with best practices @Kidwell2016-pm, and recognition of research rigour
in track record evaluations.
